---
title: "Homework 3: Alignment"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    df_print: paged
  pdf_document: default
---

```{r, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


# Instructions

- Download [this](homework.Rmd) notebook and write your solutions inside it

- Upload the notebook with your solutions to your allocated repo.

- To solve the tasks, finish the code skeletons provided.
  **The code must be runnable to get full points**.
  If you're having trouble, try to explain in your own words 
  what you were trying to do. Proper understanding is half of the points.

- Each task is independent but subtasks are linked.

- Remember to run the tests for each function you'll be completing.
And feel free to add some of your own.
By testing code in small modular parts (e.g. functions),
you can trust these parts in "downstream" (more complex) applications.

The terms "cost", "distance", and "penalty" are used interchangeably since 
in this context they represent equivalent ways of looking at the same concepts. 
It also helps illustrate that these algorithms are quite general and can be applied to many different problems.

## Advice

* Always restart R (*Session* > *Restart R*) and rerun the notebook
from the top to double-check (i.e. reproduce) your solutions.

* Remember that R starts indexing vectors at position `1`. 
Many other languages in which these algorithms are implemented start at `0`.

* If the code doesn't behave the way you want or you get errors,
try checking out the chapter on [debugging](https://adv-r.hadley.nz/debugging.html) 
from *Advanced R* for different approaches to deal with this.


# Task 1: Global Alignment

Complete the Needleman–Wunsch implementation.

## Part a: The distance between two letters [1 pt]

Fill in the "blanks" in the code below (replace the # WRITE CONDITION HERE) 
with the appropriate logical checks such that this distance function gives
the correct distance between two letters.

Remember to run the code chunk so the notebook starts using your function.

```{r}
dist_global <- function(char1, char2) {
    #############################
    # Function that computes the distance between two characters in a sequence,
    # used for global alignment.
    # 
    # Input arguments:
    #    char1, char2 = Values of 2 characters, either nucleobases or gaps, 
    #                   as 1-character strings.
    # 
    #    E.g. char1 = 'A', char2 = 'G'
    #         char1 = 'A', char2 = '-'
    # 
    # Returns:
    #    The distance (or cost of change) between char1 and char2, 
    #    using the DIST_... values
    #############################
    
    DIST_MATCH  <- 0
    DIST_GAP <- 8
    DIST_TRANSITION <- 2
    DIST_TRANSVERSION <- 4
    
    if (char1 == char2)
        return(DIST_MATCH)

    if # WRITE CONDITION HERE
        return(DIST_GAP)

    if # WRITE CONDITION HERE
        return(DIST_TRANSITION)

    if # WRITE CONDITION HERE
        return(DIST_TRANSITION)

    else
        return(DIST_TRANSVERSION)
}
```


### Tests

Check your code with the test below, assuming you use the same values as in the lecture.
If you want to experiment with different distance values, 
then naturally the test will have to be adapted.

```{r}
print(dist_global('A', 'G') == 2)
print(dist_global('A', '-') == 8)
```


## Part b: Computing the global alignment [1 pt]

Here you are provided with skeleton code to compute the global alignment
between two sequences using the Needleman–Wunsch algorithm algorithm.
The same notation is used as in the course.

```{r}
global_alignment_matrix <- function(seq1, seq2) {
    ##############################################################
    # Computes the global alignment matrix D between 
    # sequences seq1 and seq2
    # 
    # Input:
    #     seq1, seq2 = nucleobase sequences as strings
    # 
    #     E.g. seq1 = 'TATACCTGAAGGGCCT', seq2 = 'TATACGAGACCGTTT'
    # 
    # Returns:
    #     D = global alignment matrix
    ###############################################################
    
    # Convert to character vectors to easily work letter by letter
    # E.g. seq1 = "ABC" becomes c("A", "B", "C") and seq1[2] == "B"
    seq1 <- strsplit(seq1, split = "")[[1]]
    seq2 <- strsplit(seq2, split = "")[[1]]
    
    # Initialize the matrix with all zero values
    D <- matrix(0L, nrow = length(seq1) + 1, ncol = length(seq2) + 1)
    
    # Initialize horizontal (top) boundary of D using seq1
    for (i in seq(from = 2, to = length(seq1) + 1))
        D[i, 1] <-  ### WRITE CODE HERE ###
    
    # Initialize vertical (left) boundary of D using seq2
    for (j in seq(from = 2, to = length(seq2) + 1))
        D[1, j] <-  ### WRITE CODE HERE ###
    
    # Walk through all cells of the matrix,
    # updating the current cell by considering a step from 3 possible directions,
    # and taking the least "costly" of these 3 directions (shortest distance).
    for (i in seq(from = 2, to = length(seq1) + 1)) {
        for (j in seq(from = 2, to = length(seq2) + 1)) {
            dist_from_upper_left <- dist_global(seq1[i-1], seq2[j-1])
            dist_from_above      <- dist_global(seq1[i-1], "-")
            dist_from_left       <- dist_global("-",       seq2[j-1])
            
            D[i, j] <-  min(c(
                                D[i-1, j-1] +  ### WRITE CODE HERE ###  ,
                                D[i-1, j  ] +  ### WRITE CODE HERE ###  ,
                                D[i  , j-1] +  ### WRITE CODE HERE ###
                            ))
        }
    }
    return(D)
}

# This one's complete. Just a convenience function to fetch the overall value.
global_alignment_value <- function(global_alignment_matrix) {
    # Input:
    # global_alignment_matrix = a computed global alignment matrix
    # 
    # Returns:
    # The lower-right corner of the matrix, since we know that holds
    # the global alignment value
    return(global_alignment_matrix[nrow(global_alignment_matrix),
                                   ncol(global_alignment_matrix)])
}
```


### Tests

Check your code with these. The last 2 lines compare expected and actual values.

```{r}
seq1 <- "TACGTCAGC"
seq2 <- "TATGTCATGC"

expected_distance <- 10

expected_dist_matrix <-
    t(matrix(c(
        c( 0,  8, 16, 24, 32, 40, 48, 56, 64, 72, 80),
        c( 8,  0,  8, 16, 24, 32, 40, 48, 56, 64, 72),
        c(16,  8,  0,  8, 16, 24, 32, 40, 48, 56, 64),
        c(24, 16,  8,  2, 10, 18, 24, 32, 40, 48, 56),
        c(32, 24, 16, 10,  2, 10, 18, 26, 34, 40, 48),
        c(40, 32, 24, 16, 10,  2, 10, 18, 26, 34, 42),
        c(48, 40, 32, 24, 18, 10,  2, 10, 18, 26, 34),
        c(56, 48, 40, 32, 26, 18, 10,  2, 10, 18, 26),
        c(64, 56, 48, 40, 32, 26, 18, 10,  6, 10, 18),
        c(72, 64, 56, 48, 40, 34, 26, 18, 12, 10, 10)
    ), nrow = nchar(seq2) + 1, 
       ncol = nchar(seq1) + 1))


D <- global_alignment_matrix(seq1, seq2)


print(identical(D, expected_dist_matrix))
print(global_alignment_value(D) == expected_distance)
```

## ======= Solution =====


```{r}
dist_global <- function(char1, char2) {
    DIST_MATCH = 0
    DIST_GAP = 8
    DIST_TRANSITION = 2
    DIST_TRANSVERSION = 4

    if (char1 == char2)
        return(DIST_MATCH)
    
    if (char1 == '-' || char2 == '-')
        return(DIST_GAP)

    if ((char1 == 'A' && char2 == 'G') || (char1 == 'G' && char2 == 'A'))
        return(DIST_TRANSITION)

    if ((char1 == 'T' && char2 == 'C') || (char1 == 'C' && char2 == 'T'))
        return(DIST_TRANSITION)

    else
        return(DIST_TRANSVERSION)
}


global_alignment_matrix <- function(seq1, seq2) {
    # Convert to character vectors to easily work letter by letter
    # E.g. seq1 = "ABC" becomes c("A", "B", "C") and seq1[2] == "B"
    seq1 <- strsplit(seq1, split = "")[[1]]
    seq2 <- strsplit(seq2, split = "")[[1]]
    
    D <- matrix(0L, nrow = length(seq1) + 1, ncol = length(seq2) + 1)
    
    for (i in seq(from = 2, to = length(seq1)))
        D[i, 1] <- D[i - 1, 1] + dist_global(seq1[i - 1], "-")
        
    for (j in seq(from = 2, to = length(seq2)))
        D[1, j] <- D[1, j - 1] + dist_global("-", seq2[j - 1])
    
    for (i in seq(from = 2, to = length(seq1) + 1)) {
        for (j in seq(from = 2, to = length(seq2) + 1)) {
            dist_from_upper_left <- dist_global(seq1[i-1], seq2[j-1])
            dist_from_above      <- dist_global(seq1[i-1], "-")
            dist_from_left       <- dist_global("-",       seq2[j-1])
            
            D[i, j] <-  min(c(
                                D[i-1, j-1] + dist_from_upper_left,
                                D[i-1, j  ] + dist_from_above,
                                D[i  , j-1] + dist_from_left
                            ))
        }
    }
    return(D)
}

global_alignment_value <- function(global_alignment_matrix) {
    # Input:
    # global_alignment_matrix = a computed global alignment matrix
    # 
    # Returns:
    # The lower-right corner of the matrix, since we know that holds
    # the global alignment value
    return(global_alignment_matrix[nrow(global_alignment_matrix),
                                   ncol(global_alignment_matrix)])
}
```


# Task 2: Local Alignment

Complete the Smith-Waterman implementation

## Part a: The distance between two letters  [1 pt]

Finish the provided code and check yourself with the tests provided.

```{r}
dist_local <- function(char1, char2) {
    ##########################################
    # Function that computes the distance between two characters in a sequence,
    # used for local alignment.
    # 
    # Input arguments:
    #     char1, char2 = Values of 2 characters, either nucleobases or gaps,
    #                    as 1-character strings.
    # 
    #     E.g. char1 = 'A', char2 = 'G'
    #          char1 = 'A', char2 = '-'
    # 
    # Returns:
    #     The distance (or cost of change) between char1 and char2, 
    #     using the DIST_... values
    ##########################################

    DIST_MATCH <- 2
    DIST_GAP <- -6
    DIST_MISMATCH <- -4
    
    if # WRITE CONDITION HERE
        return(DIST_MATCH)
    
    if # WRITE CONDITION HERE
        return(DIST_GAP)
    
    else
        return(DIST_MISMATCH)
}
```

### Tests

```{r}
print(dist_local_example('A', '-') + dist_local_example('A', 'T') == -10)
print(dist_local_example('A', 'C') + dist_local_example('G', 'T') == -8)
```


## Part b: Computing the local alignment  [1 pt]

Finish the provided code.

Since it is a variation of the global alignment algorithm, 
the Smith-Waterman algorithm has a similar structure. 

A *big difference* is that one takes the *maximum* of different possibilities, 
since once keeps track of an optimum score.

```{r}
local_alignment_matrix <- function(seq1, seq2) {
    #######################################################
    # Computes the local alignment matrix D between 
    # sequences seq1 and seq2
    # 
    # Input:
    #    seq1, seq2 = nucleobase sequences as strings
    # 
    #    E.g. seq1 = 'TATACCTGA', seq2 = 'TATACGAGACCGTTT'
    # 
    # Returns:
    #    D = local alignment matrix
    ########################################################
    # Convert to character vectors to easily work letter by letter
    # E.g. seq1 = "ABC" becomes c("A", "B", "C") and seq1[2] == "B"
    seq1 <- strsplit(seq1, split = "")[[1]]
    seq2 <- strsplit(seq2, split = "")[[1]]    

    # Initialize the matrix with all zero values
    D <- matrix(0L, nrow = length(seq1) + 1, ncol = length(seq2) + 1)
    
    # Walk through all cells of the matrix,
    # updating the current cell by considering a step from 3 possible directions,
    # and taking the max of scores corresponding 3 directions.
    for (i in seq(from = 2, to = length(seq1))) {
        for (j in seq(from = 2, to = length(seq2))) {
            
            dist_from_upper_left <- ### WRITE CODE HERE ###
            dist_from_above      <- ### WRITE CODE HERE ###
            dist_from_left       <- ### WRITE CODE HERE ###
            
            D[i, j] = max(c(  ### WRITE CODE HERE ###  ,
                              ### WRITE CODE HERE ###  ,
                              ### WRITE CODE HERE ###  ,
                          0)) # make sure to have a non-negative value since this is a distance
        }
    }
    return(D)
}


local_alignment_value <- function(alignment_matrix) {
    max(alignment_matrix)
}
```

### Tests

Check your code with these. The last 2 lines compare expected and actual values.

```{r}
seq1 <- "GGTATGCTGGCGCTA"
seq2 <- "TATATGCGGCGTTT"

expected_distance <- 12

expected_dist_matrix <- 
    t(matrix(c(
       c( 0,  0,  0,  0,  0,  0,  0,  0,  0,  0,  0,  0,  0,  0,  0),
       c( 0,  0,  0,  0,  0,  0,  2,  0,  2,  2,  0,  2,  0,  0,  0),
       c( 0,  0,  0,  0,  0,  0,  2,  0,  2,  4,  0,  2,  0,  0,  0),
       c( 0,  2,  0,  2,  0,  2,  0,  0,  0,  0,  0,  0,  4,  2,  2),
       c( 0,  0,  4,  0,  4,  0,  0,  0,  0,  0,  0,  0,  0,  0,  0),
       c( 0,  2,  0,  6,  0,  6,  0,  0,  0,  0,  0,  0,  2,  2,  2),
       c( 0,  0,  0,  0,  2,  0,  8,  2,  2,  2,  0,  2,  0,  0,  0),
       c( 0,  0,  0,  0,  0,  0,  2, 10,  4,  0,  4,  0,  0,  0,  0),
       c( 0,  2,  0,  2,  0,  2,  0,  4,  6,  0,  0,  0,  2,  2,  2),
       c( 0,  0,  0,  0,  0,  0,  4,  0,  6,  8,  2,  2,  0,  0,  0),
       c( 0,  0,  0,  0,  0,  0,  2,  0,  2,  8,  4,  4,  0,  0,  0),
       c( 0,  0,  0,  0,  0,  0,  0,  4,  0,  2, 10,  4,  0,  0,  0),
       c( 0,  0,  0,  0,  0,  0,  2,  0,  6,  2,  4, 12,  6,  0,  0),
       c( 0,  0,  0,  0,  0,  0,  0,  4,  0,  2,  4,  6,  8,  2,  0),
       c( 0,  2,  0,  2,  0,  2,  0,  0,  0,  0,  0,  0,  8, 10,  4),
       c( 0,  0,  4,  0,  4,  0,  0,  0,  0,  0,  0,  0,  2,  4,  6)
    ), nrow = nchar(seq2) + 1, 
       ncol = nchar(seq1) + 1))


D <- local_alignment_matrix(seq1, seq2)


print(identical(D, expected_dist_matrix))
print(local_alignment_value(D) == expected_distance)
```


## ======= Solution =====

```{r}
dist_local <- function(char1, char2) {
    DIST_MATCH <- 2
    DIST_GAP <- -6
    DIST_MISMATCH <- -4
    
    if (char1 == char2)
        return(DIST_MATCH)
    
    if (char1 == '-' || char2 == '-')
        return(DIST_GAP)
    
    else
        return(DIST_MISMATCH)
}

local_alignment_matrix <- function(seq1, seq2) {
    seq1 <- strsplit(seq1, split = "")[[1]]
    seq2 <- strsplit(seq2, split = "")[[1]]    

    D <- matrix(0L, nrow = length(seq1) + 1, ncol = length(seq2) + 1)
    
    for (i in seq(from = 2, to = length(seq1) + 1)) {
        for (j in seq(from = 2, to = length(seq2) + 1)) {
            
            dist_from_upper_left <- dist_local(seq1[i - 1], seq2[j - 1])
            dist_from_above      <- dist_local(seq1[i - 1], '-')
            dist_from_left       <- dist_local('-', seq2[j - 1])
            
            D[i, j] = max(c(D[i - 1, j - 1] + dist_from_upper_left,
                            D[i - 1, j] + dist_from_above,
                            D[i, j - 1] + dist_from_left,
                          0)) # make sure to have a non-negative value since this is a distance
        }
    }
    return(D)
}


local_alignment_value <- function(alignment_matrix) {
    max(alignment_matrix)
}
```



# Task 3: Assembly

Finish the code below to compute the **shortest common superstring** (SCS)
using the **greedy algorithm**.


## Part a: Max overlap of two sequences  [1 pt]

To assemble reads, we need a function to compute the length of the overlap
of two given sequences

For example:

```
seq1: abc123
seq2:    123xyz

len_overlap('abc123', '123xyz') == 3
```

The skeleton for this function is given below. Finish the code and use the tests
below to check yourself.

Note that this time we're not converting the strings `seq1` and `seq2` to character vectors, 
because R has some functions which we can use, but these work on strings 
(e.g. "hello" instead of c("h", "e", "l", "l", "o")).
Read about these functions: `substr()`, `startsWith()`, `endsWith()` and think
about how to use them to solve this tasks. 
Also, recall `nchar()` counts the number of characters in a string, while `length()`
counts the elements in a vector. A string is a single element, so `length("hello") == 1`
and `nchar("hello") == 5`.

You are of course free to expand the function if you feel constrained by the given
"blanks". You can even rewrite it entirely as you wish, just please explain what you're doing.

**! Careful !**

The R `while(CONDITION)` loop command simply repeats until CONDITION is no longer true. 
If there is a mistake in the code and CONDITION never changes,
the code will run forever. If you see your code running for very long 
(like when running the tests), press the red stop button in the corner of the code chunk. 
Sometimes RStudio doesn't like this.
If it starts misbehaving (like not running your code chunks), try restarting R
(Session > Restart R) or getting a new RStudio session (Session > New Session).

```{r}
len_overlap <- function(seq1, seq2) {
    ##############################################    
    # Find the longest overlap between a suffix of seq1 and a prefix of seq2
    # and return its length.
    # E.g. len_overlap('abc123', '123xyz') == 3
    ##############################################

    # We take shorter and shorter suffixes of seq1
    # and test whether they overlap with a prefix of seq2.
    # We start at index 1 in seq1 (i.e. thus including seq1 entirely)
    overlap_start_index <- 1
    while(overlap_start_index <= ## WRITE CODE HERE ##  ) {
          
        current_prefix <- ## WRITE CODE HERE ##
        
        if (## WRITE CODE HERE ##   ) {
            return(nchar(seq1) - overlap_start_index + 1)
        }
        else
            overlap_start_index = overlap_start_index + 1
    }
    return(0)  # no overlap
)
```

### Tests

You can add your own of course. Think of edge cases and negative examples.

```{r}
print(len_overlap('abc123', '123xyz') == 3)
print(len_overlap('abc1234', '1234xyz') == 4)
print(len_overlap('abc124', '123xyz') == 0)
```


### ======== Solution =========

```{r}
len_overlap <- function(seq1, seq2) {
    overlap_start_index <- 1
    while(overlap_start_index <= nchar(seq1)) {
        current_prefix <- substr(seq1, start = overlap_start_index, stop = nchar(seq1))
        if (startsWith(x = seq2, prefix = current_prefix)) {
            return(nchar(seq1) - overlap_start_index + 1)
        }
        else
            overlap_start_index = overlap_start_index + 1
    }
    return(0)
}

print(len_overlap('abc122', '22xyz') == 2)
print(len_overlap('abc124', '123xyz') == 0)
print(len_overlap('abc1234', '1234xyz') == 4)
print(len_overlap('abc1234', '1234xyzuvw') == 4)
print(len_overlap('1234', '1234xyz') == 4)
print(len_overlap('abc1234', '1234') == 4)
print(len_overlap('abc1234', '4123xyz') == 1)
```


## Part b: Greedy Shortest Common Superstring  [1 pt]

This approach requires generating all permutations of a vector of strings.
To make our lives easier, we're going to use the `permn()` function
from the `combinat` package. So first install this package:

```{r}
if (!require("combinat"))
    install.packages("combinat")
```

To see how it works, run the chunk below:

```{r}
library(combinat)

test_permutations <- combinat::permn(c("I", "prefer", "Python"))
test_permutations
```

Remember there are `n!` (n factorial) permutations of `n` elements, so we should
refrain from long vectors (e.g. `factorial(10)` is already `3628800`).

```{r}
library(combinat)

greedy_shortest_common_superstring <- function(strings) {
    #############################################################    
    # Takes a list of strings and computes
    # an approximation of the shortest common supperstring (SCS),
    # using a greedy algorithm.
    # 
    # Input:
    #       strings = list of strings e.g. c('ABCD', 'CDBC', 'BCDA')
    # 
    # Returns:
    #       shortest_superstring = the approximation of the SCS 
    #                              of the input list of strings
    # E.g. 'ABCDBCDA'
    ###############################################################
    
    # Start off with no SCS (empty string)
    shortest_superstring <- ""
    all_permutations_of_the_strings <- combinat::permn(strings)
    
    # Consider all permutations of the input list of stings
    for (current_permutation in all_permutations_of_the_strings) {

        # For the current permutation,
        # start growing a superstring using the first string in the permutation
        superstring_candidate <- current_permutation[1]
        
        # Go over all strings in this permutation and overlap consecutive strings i and i + 1
        # So e.g. i: ABB   i+1: BBA  ->  ABBA
        for (i in seq(from = 1, to = length(strings) - 1)) {
            # determine length of the overlap ...
            overlap_length <- len_overlap(#### WRITE CODE HERE ####  )

            # ... and use this to get the non-overlapping part of string i + 1
            non_overlapping_portion <- #### WRITE CODE HERE ####

            # The non-overlapping part of string i + 1 is then pasted 
            # to the end of the superstring we are growing for this permutation
            superstring_candidate <- paste0(superstring_candidate, non_overlapping_portion)
        }
    
        # If this is the first permutation we're considering,
        # of it turns out that the superstring candidate we grew for this permutation of input strings
        # is shorter than our current SCS candidate, 
        # make the current superstring candidate the new SCS
        if (shortest_superstring == "" || nchar(superstring_candidate) < #### WRITE CODE HERE ####  )
            shortest_superstring <- superstring_candidate
    }
    return(shortest_superstring)
}
```


### Tests

The second case will be a bit slow, because of the many permutations that are considered.
You can add your own tests of course. Think of edge cases.

Note that there may be multiple SCS and the procedure will choose only the first one,
and the order is determined by how the permutations are generated. 
So if you add your own test, check the output of `permn(my_test_vector)` to see the order
in which each SCS candidate is constructed.

```{r}
print(greedy_shortest_common_superstring(c('ABCD', 'CDBC', 'BCDA')) == 'ABCDBCDA')
print(greedy_shortest_common_superstring(c('BAA', 'AAB', 'BBA', 'ABA', 'ABB', 'BBB', 'AAA', 'BAB') == 'BAAABABBBA'))
```



## ======= Solution ======

```{r}
library(combinat)

greedy_shortest_common_superstring <- function(strings) {
    #############################################################    
    # Takes a list of strings and computes
    # an approximation of the shortest common supperstring (SCS),
    # using a greedy algorithm.
    # 
    # Input:
    #       strings = list of strings e.g. c('ABCD', 'CDBC', 'BCDA')
    # 
    # Returns:
    #       shortest_superstring = the approximation of the SCS 
    #                              of the input list of strings
    # E.g. 'ABCDBCDA'
    ###############################################################
    
    # Start off with no SCS (empty string)
    shortest_superstring <- ""
    all_permutations_of_the_strings <- combinat::permn(strings)
    
    # Consider all permutations of the input list of stings
    for (current_permutation in all_permutations_of_the_strings) {

        # For the current permutation,
        # start growing a superstring using the first string in the permutation
        superstring_candidate <- current_permutation[1]
        
        # Go over all strings in this permutation and overlap consecutive strings i and i + 1
        # So e.g. i: ABB   i+1: BBA  ->  ABBA
        for (i in seq(from = 1, to = length(strings) - 1)) {
            # determine length of the overlap ...
            overlap_length <- len_overlap(current_permutation[i], 
                                          current_permutation[i + 1])

            # ... and use this to get the non-overlapping part of string i + 1
            non_overlapping_portion <- substr(current_permutation[i + 1],
                                              start = overlap_length + 1,
                                              stop = nchar(current_permutation[i + 1]))

            # The non-overlapping part of string i + 1 is then pasted 
            # to the end of the superstring we are growing for this permutation
            superstring_candidate <- paste0(superstring_candidate, non_overlapping_portion)
        }
    
        # If this is the first permutation we're considering,
        # of it turns out that the superstring candidate we grew for this permutation of input strings
        # is shorter than our current SCS candidate, 
        # make the current superstring candidate the new SCS
        if (shortest_superstring == "" || nchar(superstring_candidate) < nchar(shortest_superstring))
            shortest_superstring <- superstring_candidate
    }
    return(shortest_superstring)
}

print(greedy_shortest_common_superstring(c('ABCD', 'CDBC', 'BCDA')) == 'ABCDBCDA')
print(greedy_shortest_common_superstring(c('BAA', 'AAB', 'BBA', 'ABA', 'ABB', 'BBB', 'AAA', 'BAB')) == 'BAAABBBABA')
```


