---
title: "Homework 4: Hidden Markov Models"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    df_print: paged
  pdf_document: default
---

# Instructions

For this homework, your task is to build an HMM model for annotating S. cerevisiae. Please read through the instructions before you start, to best plan your work.

The work is split up in four sections:

0. Setting up your directories and getting your data 
1. getting baseline results with the HMM-based tool Augustus (code provided)
2. reading a little reference material on HMM gene prediction
3. Going through a toy example that will show you how to:
    * Initiate an HMM model in the R package HMM
    * Use the defined helper functions to extract data sets and  
    * Train a model on data 
    * compare a trained model with base line results from augustus
4. building and training your own HMM

**To complete this homework you need to:**

    1. Initiate your own HMM model
        * Explain the architecture of your model states emmissions etc.
        * Explain what initial parameters you chose and why.
        * Explain if you use any of the references provided or you found your own.
        * Explain if you did test different model architectures or different initial parameters (Not a requirement that you do)
    2. Train your model on data
        * Explain what data you used as training data and why (what chromosomes)
        * Explain what training algorithm you used, did you test both?
        * Explain how many training iterations did you use and why that number.
    3. Evaluate your model with augustus results as shown in the example code.
        * Explain what data you used as testing data (should not be the same as training data).
        * Show at least three histograms with the augustus results initial model and trained model
        * Discuss shortly the evaluation, did your model improve with training, did you get close to augustus?
    4. Given unlimited resorces in both compute and time would you have liked test something more? (just a few sentences)
    
You can work in this notebook and put in your own code. The answers you can type in at the very bottom or in a separate .txt file named Answers-HW4 The histograms can still be in the notebook just specify what line they are generated at. 

This homework together with the project is meant for you to feel a little of how it is like to work as a bioinformatician. Thus, for this homework feel free to experiment and change things if you think there is a better way to do things another way. Just specify what you changed and why. There is of course obvious ways to improve validation and also the annotation of the genes with different patterns. **Most importantly this is meant to be fun!**  

# 0. Setting up your directory
Start by making a HW4 directory. In your HW4 directory you will create a results directory and you will create a data directory. In your data directory you'll place a copy of the file S288C_reference_sequence_R64-2-1_20150113.fsa that we used in the Unix HW. This notebook should be placed in the HW4 directory. After that we should be god to go.  

# 1. Baseline Results
Note This step can take some time (~30 min if the server isn't under heavy load).
For this part, you will need to run Augustus on the server, where it's already installed.
Grab a copy of the S. cerevisiae reference genome as described in section 0. Then run Augustus as below (adjust if you've named your directories differently)

```{bash, eval=F}
augustus data/S288C_reference_sequence_R64-2-1_20150113.fsa \
    --species=saccharomyces_cerevisiae_S288C \
    --progress=True > results/ref_annot_augustus.gff

```
to view the structure of the data in the .gff file run the code block below.
```{bash, eval=F}

sed '1,14d' results/ref_annot_augustus.gff | head -n 12

```

Augustus basically lists all found genes one after the other, with the various sections marked. It also gives the protein sequence as a comment. Columns to note are:

1. seqID = chromosome ID
2. type of feature
3. start position of feature
4. end position of feature
  
Extract the protein sequences into a FASTA file. This will be used to compare your results against later in the evaluation step. Note: the script getAnnoFast.pl is located in directory /home/sandra/Documents/HMM/
And it uses the .gff file created by augustus and places the result in a seperate file called `ref_annot_augustus.aa`

```{bash, eval=F}
/home/sandra/Documents/HMM/getAnnoFast.pl results/ref_annot_augustus.gff
```

# 2. Literature

Read at least the 2 references below (a few pages), to recap the concepts and to maybe get a model as your starting point.
Please very briefly describe your chosen model. Why did you or the authors chose the sates you did? Did the authors mention why they chose the probabilities they did? If you created your own model, how did you decide the probabilities?

References:

- Quick recap of gene annotation with HMMs, along with a simple example model [What is a hidden Markov model?](https://www.nature.com/articles/nbt1004-1315)

- Some lecture slides from the University of Waterloo (pages 1-14 cover models like we've discussed). You could use the model on page 9 as a starting point https://www.math.uwaterloo.ca/~aghodsib/courses/w05stat440/w05stat440-notes/feb27.pdf

- Yet another article describing a number of HMM classes for sequence analysis The model in 2.2. (Fig. 1) could also be a starting point https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2766791/


# 3. The basic (2-state) model 

We'll be playing with simple HMMs using the HMM package, the documentation can be found here [HMM](https://cran.r-project.org/web/packages/HMM/HMM.pdf).

Install the necessary packages first.
```{r,  message=F, warning=F, eval=F}
# Making sure all packages that we need is installed
if (!require("HMM"))
    install.packages("HMM")
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if(!require("Biostrings"))
    BiocManager::install("Biostrings")
```

Import the packages.

```{r, eval=F}
# Importing the stuff we need. The message and warning flags above are to get RStudio to shut up
# when importing these libraries (not print all the messages dumped by them).
library("HMM")
library("Biostrings")
library("tidyverse")
```

### Defining some helper functions
You will use these functions when you set up your own experiments. 

1. `preparing_observations(set, indices)` helps you prepare your data sets 
    * set: expects a DNAStringSet (your loaded reference genome)
    * indices: expect indices of which chromosomes to load (1-17)
    * returns: list of observations
2. `get_gene_start_and_stop_from_annotation(annotation_list, pattern)` helps you get the indices of annotated genes
    * annotation_list: expects annotated observations as a list.
    * pattern: expects regex string (use default 'C+' to get all starts and stops of stretches of C)
    * returns: list of start and stop indices for annotated genes (as pattern describes)
3. `convert_to_proteins(indices, string)` helps you to translate annotated genes to proteins
    * indices: expect output from get_gene_start_stop (first half of list start indices second half stop indices)
    * string: expects DNAString to be translated
    * returns: list of proteins 
    
```{r, eval=F}
get_gene_start_and_stop_from_annotation <- function(annotation_list, pattern){
  # Function that returns indices of all matches to the pattern
  annotation_string <- convert_list_to_string(annotation_list)
  list_of_indices <- stringr::str_locate_all(annotation_string, stringr::regex(pattern))
  return (list_of_indices)
}

convert_list_to_string <- function(list) {
  #I'm just a helper function that converts lists of individual letters into DNA strings 
  #. Please ignore me.
  return(paste(list, collapse=''))
}
convert_string_to_list = function(string) {
    # I'm just a helper function that converts DNA strings into lists of individual
    # letters so the HMM functions don't complain. Please ignore me.
    return(stringr::str_split(string, pattern = "")[[1]])
}

convert_to_protein <- function(start, stop, string=""){
  # convert a DNAstring to a AAsequence between the start and stop indices
  return(Biostrings::translate(string[start:stop]))
}

convert_to_proteins = function(indices, string){
  # Function that converts the DNA string in to proteins where the indices indicate the coding regions are.
  string <- convert_list_to_string(string)
  indices <- unlist(indices)
  n_indices = length(indices)/2
  start <- indices[1:n_indices]
  stop  <- indices[(n_indices+1):(n_indices*2)]
  DNA_string <- Biostrings::DNAString(string)
  proteins <- mapply(convert_to_protein, start, stop, MoreArgs = list(string=DNA_string), SIMPLIFY = TRUE )
  return(AAStringSet(proteins))
}

preparing_observations <- function(set, indices){
  # prepare data by concatenating all the DNAstrings in set specified by indices 
  observations <- purrr::flatten_chr(sapply(set[indices], convert_string_to_list))
  return (observations)
}
```

### Initiate HMM model

Now it's time to initiate our model, we can do that with the function `initHMM()` which is part of the HMM module.
We need to specify starting values for the start/transition/emission probabilities as well as the state/emission symbols. 
Below is the same basic model we've seen in the lecture and exercise, provided to show how to use the package.
Skim through the short hmm tutorial to fix some concepts if the code/comments below are confusing.

```{r, eval=F}
# initiates simple HMM model
gene_annot_model = initHMM(
    States = c("C", "N"),
    Symbols = c("A", "T", "C", "G"),       # Symbols: set of outputs Y  (the order here determines the order of values below!)
    startProbs = c(0.3, 0.7),              # startProbs: initial transition probabilities $\pi$ into each state
    transProbs = rbind(c(0.995, 0.005),    # transProbs: the matrix A: each row c(...) holds the outgoing transition probabs for a given state (order),
                       c(0.01,  0.99)),    #             rbind() = "row bind": creates a matrix, which is the strucure this functions expects
    emissionProbs = t(rbind(c(0.3, 0.1),   # emissionProbs: the matrix B: each column represents the $b_state(symbol)$ for a given state.
                            c(0.3, 0.1),   #                t() is matrix transpose (because the function wants $b_state(symbol)$ to be rows
                            c(0.2, 0.4),   #                but the notes use columns...)
                            c(0.2, 0.4)))
)
print(gene_annot_model)  # Nice description of the model to inspect what we did.
```

### Prepare Data

To train our HMM model the HMM module expects the observations to be in a sequence form, we can do this by using the helper function `preparing_observations(set, indices)` and give it the arguments set and indices. Set is the DNAstring set (reference genome) and indices is a vector specifying which chromosomes to load.

When training a model it is **important to validate your improved model on data that the model has not yet seen**. Thus we will prepare **two** data sets one for **training** and one for **testing**. We do this separation by choosing different chromosomes for the two different sets. in the code below we have chosen chromosome 1-2 for the training set and chromosome 3-4 for the test set.

```{r, eval=F}
# load reference genome
reference_genome_set <- readDNAStringSet("data/S288C_reference_sequence_R64-2-1_20150113.fsa")

#preparing training set using chromosome 1-2  
training_observations = preparing_observations(reference_genome_set, c(1:2))
#preparing test set using chromosomes 3-4
test_observations = preparing_observations(reference_genome_set, c(3:4))
```

### Training model

Now that we have initialized our model and created our training data we are ready to start training the model. We do that by either the `baumWelch(HMM_model, observations, maxIterations)` or `viterbiTraining(HMM_model, observations, maxIterations)` which is the two algorithms for training in the HMM module. The argument `HMM_model` is your HMM model, `observations` is the training data and `maxIterations` is the maximum number of iterations that the algorithm will run for (there is a second stopping condition that has to do with a convergence criteria). In the code example below we train our model for 10 iterations, which can take a few minutes.

Notes: 

1. We train our model on training data.
2. In this example we have wraped the training expression `(trained_model <- viterbiTraining(gene_annot_model, training_observations, maxIterations=10)[[1]])` in `system.time()` to get the running time for the training. 

```{r, eval=F}
system.time(trained_model <- viterbiTraining(gene_annot_model, training_observations, maxIterations=10)[[1]])

```

### Saving model and loading model

We dont have this as part of this homework but it could be convinent for you to save parameter values once models are trained (it can take a long time to train models). Maybe you would like to return to a model for further testing or you are supper happy with it and wants to save it to use it later. You can read up on saving and loading basic data object in R on [saving data](https://rstudio-education.github.io/hopr/dataio.html)

### Annotate chromosome

Once the model is trained we can use the trained model to annotate our test chromosome with `C` or `N` (coding or non-coding nucleobase). We do this with the function `viterbi(HMM_model, observations)`, below we annotate the test set both using the initial model and the model that we trained. 

After annotating we can gather the indices for the predicted coding regions of the test data. With the indices of the coding region and the DNA sequence we can translate the regions in to their respective proteins with the helper function `convert_to_proteins(indices, observations)`.  

```{r, eval=F}
PATTERN <- "C+" # Simple pattern getting the 
annotation_initial_model <- viterbi(gene_annot_model, test_observations)
annotation_trained_model <- viterbi(trained_model, test_observations)
annotation_indices_initial_model <- get_gene_start_and_stop_from_annotation(annotation_initial_model, PATTERN)
annotation_indices_trained_model <- get_gene_start_and_stop_from_annotation(annotation_trained_model, PATTERN)
proteins_initial_model <- convert_to_proteins(annotation_indices_initial_model, test_observations)
proteins_trained_model <- convert_to_proteins(annotation_indices_trained_model, test_observations)
```


### Comparing gene length distributions  to reference

Now that we have translated our coding regions we can compare the length distributions of our proteins from the initial and trained models with the proteins that Augustus found. This way we can see if we are improving with training. 

```{r, eval=F}
width_augustus <- width(readAAStringSet("results/ref_annot_augustus.aa"))
width_initial_model <- width(proteins_initial_model)
width_trained_model <- width(proteins_trained_model)
hist(width_augustus)
hist(width_initial_model)
hist(width_trained_model)
```

### Some caveats

1: The Augustus distribution is from the whole genome not just the training set, if you want to you can save the specific chromosomes that you have in your test set in a separate .fasta file and re run augustus on that to only extract the proteins corresponding to chromosomes in your test set. In a real world scenario that would be the way to do it. However we don't expect to much differences in the length distribution between chromosomes.  

2: There are other methods for measuring the performance of your model, you could for example download the proteome from a data base like NCBI instead of using Augustus. You could also compare the proteins predicted by your model and the reference proteome using MSA to see the differences, for our purposes that is over kill. 

3: Here we used a simple model with only two states coding and non-coding however DNA, as you know, have many more distinct regions with different roles. most often in machine learning picking the right model architecture that best describes the data is an art and well worth spending some time on.

4: When we initiated our models parameters we already knew roughly what they should be based on our knowledge of biology. However, it is not always the case that we have such prior knowledge, in those cases we usually start from a uniform distribution of the parameters.

5: When translating our predicted genes you can see that some places is annotated with * rather then a character of an amino acid these residues indicate instead that a stop codon was translated. 

6: The function provided for you to extract gene positions requires a regular expression (pattern) that describes what a gene looks like given the hidden states you chose ("C+" here). If you have trouble adjusting it for your model, get in touch with us since regexes were not covered in the course. If you want to experiment, https://regex101.com/ is a great size that tests and analyzes regexes

# 4. Assignment: Build a better HMM

Your turn! Create a better HMM using the example and code provided and try your best to get better results.
How well can you do against Augustus? :) (the important thing is to build + use the model so don't worry about matching it - just aim to improve results)


Remember to Have fun!
