---
title: "Homework 2: Intro to R programming for data analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    df_print: paged
  pdf_document: default
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Instructions

Create an R notebook and write the R code that solves the tasks below.
Then add this notebook to your **homework repo** 
(make sure you actually **push to GitHub**). 
Try to work on the server. That's where the data files are located.

To **pass the homework** you need **5 points**. 
The tasks are meant to be rather independent (and a few have subtasks).

**To get full points**, please:

1. **`[1 pt]`** Describe what the code is doing. Even if you find solutions online, 
it's important that you understand what's going on.

2. **`[1 pt]`** **Make sure your notebook can be run from top to bottom** (i.e. is reproducible). 
To check, restart R (Session menu > Restart R) to get a fresh session,
then click on the Run button (upper-right) > Run All. 
This is **crucial** because as you work, various variables will accumulate in the session and this can "sabotage" your code.

**Bonus point**
Write code that is understandable :) I won't mind not choosing the best possible names 
for variables but try to inform the reader what's going on by naming things in a clear way. 
So `reference_genome` rather than `df` or `x`. This is also a service to future you! 


## Advice

- Remember to always inspect both the input data, as well as intermediate results
- We try to point to relevant sources of information, but using the Internet and
biological databases is a skill in itself, so if the contents of certain biological files
are confusing, try to find information about the formats. 
Most databases have these described somewhere.



# Task 1: How many genes are on yeast S288C chromosome II? [1 pt]

Let's redo task 1 in homework 1 using R and tidyverse.
Read in `/home/fburic/data/S288C_reference_genome_R64-2-1_20150113/saccharomyces_cerevisiae_R64-2-1_20150113.gff`
as a table (Hint: use the `read_tsv()` function). 
Note that *the first column is the chromosome*, not the sequence ID as in other GFF files.

Count the number of genes on chromosome II using table manipulation (tidyverse)
functions like you've seen in the tutorial.
Please have your code produce a numerical result.
The interactive RStudio display of the number of rows doesn't count :)

If you require other functions than what we've covered, take look at this
[cheat sheet](https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf).



# Task 2: Compute average gene length by chromosome in the S288C yeast strain  [1 pts]

Using the same table you used in the Task 1, calculate the average gene length
**per chromosome**. Remember to inspect your data, to see what you're working with.
Please use the table (tidyverse) manipulation functions we've covered in the tutorial.



# Task 3: Extract gene IDs from the S288C yeast GFF file and look for some genes [2 pts]

Using the S288C GFF file again, you may have noticed the attributes column contains
a lot of information, some of which is quite important, like the gene ID.

Here's a function that extracts the gene ID from that column:

```{r}
extract_gene_id_from_attributes_s288c <- function(attribute_string) {
    # str_split heard we like data structs 
    # so it put lists in our lists so we can index while we index
    id_field <- stringr::str_split(attribute_string, pattern = ";")[[1]][[1]]
    actual_id <- stringr::str_split(id_field, pattern = "=")[[1]][[2]]
    return(actual_id)
}
```

a. Using the table manipulation functions, create a new column `gene_id` in the
GFF table. You'll find the pattern in the last exercise of the tutorial.
Just use `sapply` instead. [1 pt]

b. Write code (using the table manipulation functions) that 1) checks whether the following
gene IDs are present in the table and 2) only prints the chromosomes on which these are found. [1 pt]
Hint: For 1), check out the `%in%` function though it's not the only way.
(Obs: only 0.5 points if you have a separate copy of the code for each gene ID :) )

```
YJR117W
YKL156C-A
YOL110W
YJR104C
YJL111W
YXYZZY
YNOPE
```



# Task 4: Find top 5 longest genes in the S288C yeast  [1 pt]

Compute gene lengths as before, but this time simply get the highest 5 values.
There are several ways to do this using the basic table manipulation functions you've seen
in the tutorial. If you're having trouble with those, 
maybe there's something else you could use from the  
[cheat sheet](https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf).




# Task 5: Calculations per gene in the Y55 strain [2 pts]

The yeast Y55 strain FASTA file is here: `/home/fburic/data/Y55_JRIF00000000_SGD_cds.fsa`

As a demo of Biostrings, let's compute the GC content per gene.
To do this you first need to load in the `.fsa` file with the Biostrings function 
`readDNAStringSet()` just as shown during the tutorial 
(in a variable called e.g. `genes_y55_as_stringset`).

Write that small bit of code to load the `.fsa` file, 
since you'll need it, plus the code below to solve this task.

To compute the GC contents you can use a Biostrings function called `letterFrequency()`
You can run the command directly on the loaded data like this:

```{r}
library(Biostrings)
gc_content_y55 <- Biostrings::letterFrequency(genes_y55_as_stringset,
                                              letters = c("GC"), 
                                              as.prob = TRUE)

head(gc_content_y55)   # only show the first few genes
```
Biostrings has a bunch of built-in functions for typical calculations like GC-content.
See here for examples:

- https://monashbioinformaticsplatform.github.io/r-more/topics/sequences_and_features.html
- https://web.stanford.edu/class/bios221/labs/biostrings/lab_1_biostrings.html

But what happens if we want to compute something else not supported by Biostrings?
Then the `DNAStringsSet` structure can be difficult to work with.

So let's convert it to a data frame instead using the `data.frame()`
function, setting the columns as `gene_info` and `sequence`:

```{r}
genes_y55 <- data.frame(gene_info = names(genes_y55_as_stringset), 
                        sequence = paste(genes_y55_as_stringset))
```

Here's a function to extract the gene ID from the `gene_info` fields.


```{r}
extract_gene_id_from_names_y55 <- function(name) {
    stringr::str_split(name, pattern = "_")[[1]][[1]]
}
```

a. Use the function to add a `gene_id` column. 
You'll find the pattern in the last exercise of the tutorial.
Just use `sapply` instead. [1 pt]

Hint for cleaning up: you can delete a column by "deselecting" it:
`dataframe %>% select(-column_to_delete)`

b. Rare codons are biologically significant
(here's an [example](https://www.nature.com/articles/srep20608), if you're curious).
In yeast, the rarest codon (besides stop codons) is `CGG` (mRNA) [[source](https://www.kazusa.or.jp/codon/cgi-bin/showcodon.cgi?species=4932)].
So, using table manipulation functions on the resulting table from a.,
create a column to count the occurrence of `GCC` (DNA) per gene.
Hint: take a look at what functions [stringr](https://stringr.tidyverse.org/) has to offer.




# Task 6: Search for a motif in all genes [1 pt]

Filter for those genes that contain the sequence motif "CCACACCACACCCAC".
Read the Biostrings documentation or Google what function to use (hint: it's a single function)

So your code should be something like:

```r
motif <- "CCACACCACACCCAC"
motif_hits <- Biostrings:: < find the function > (motif, reference_genome_as_stringset)
```

Then convert the hits into a data frame as `as.data.frame(motif_hits)`.
The column "group" means "chromosome" since, recall, the input FASTA file simply has
full chromosomes as entries, not individual genes. ("group_name" is irrelevant here.)

Which chromosome has the most hits? 
Write the code solution using table manipulation functions.
Your code should produce only the chromosome number with the highest motif count
(i.e. not just printing the table of all chromosomes).


